Prediction about 2024 U.S. Presidential Election Outcome Using Bayesian Modeling

### 2024-11-01 paper/paper.qmd
Updated discussion and model details

The remainder of this paper is structured as follows. @sec-data introduces the overview of the data (@sec-data-overview), measurement (@sec-data-measure), as well as explanations of outcome (@sec-data-outcome) and predictor variables (@sec-data-predictor). @sec-model explains the modeling process in detail, including procedure of model set-up (@sec-model-set), model justification (@sec-model-justify) and model validation (@sec-model-validate). Then, @sec-result presents the prediction outcome, and @sec-discussion discusses the results and models. Appendix involves three parts: [Appendix -@sec-appx-pollster] presents methodology overview and evaluation of one certain pollster: Emerson; [Appendix -@sec-appx-survey] provides detailed idealized methodology and survey design for the poll; [Appendix -@sec-model-details] presents additional details during modeling process.

@@

## Model set-up {#sec-model-set}

@@

### Bayesian model with spline for vote percentage nationwide

@@

### Bayesian model with spline for vote percentage and state as fixed effect

@@

### Assumptions of the Bayesian models

@@

## Model justification {#sec-model-justify}

@@

## Model Validation {#sec-model-validate}

During the modeling process, posterior predictive checks (PPC) is used. PPC is the comparison between what the fitted model predicts and the actual observed data, which validates whether the fitted model is compatible with the observed data. The aim is to detect if the model is inadequate to describe the data [@ppcheck].

@fig-model1ppcheck shows the posterior predictive checks for the first model, which predicts Trump's percentage vote over time in the national perspective.

@@

```{r}
#| echo: false
#| eval: true
#| label: fig-model1ppcheck
#| fig-cap: "Posterior predictive checks for model 1 (national)"
#| warning: false

pp_check(spline_model_national)
```

In @fig-model1ppcheck, the black line basically follows the trend of the light blue lines, indicating the observed data consistently falling within the range of simulated data. This suggests that the model provides a good fit, effectively predicting the actual trend of the data.

Similarly, @fig-model2ppcheck shows the posterior predictive checks for the second model, which predicts Trump's percentage vote over time by state.

```{r}
#| echo: false
#| eval: true
#| label: fig-model2ppcheck
#| fig-cap: "Posterior predictive checks for model 2 (by state)"
#| warning: false

In @fig-model2ppcheck, Although the black line has minor deviation away from the light blue lines, and a part of the black line falls outside the range of simulated lines, the majority part of the black line matches with the light blue lines.

@@

## Predictions and spline fit for vote percentage nationally

@@

## Prediction and spline fit for vote percentage by state

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

@@

@@

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

#pp_check(first_model) +
#  theme_classic() +
#  theme(legend.position = "bottom")

#posterior_vs_prior(first_model) +
#  theme_minimal() +
#  scale_color_brewer(palette = "Set1") +
#  theme(legend.position = "bottom") +
#  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

#plot(first_model, "trace")

#plot(first_model, "rhat")
```

>>>>

The remainder of this paper is structured as follows:

@sec-data introduces the overview of the data (@sec-data-overview), measurement (@sec-data-measure), as well as explanations of outcome (@sec-data-outcome) and predictor variables (@sec-data-predictor). @sec-model explains the modeling process in detail, including procedure of model set-up (@sec-model-set) and model justification (@sec-model-justify). Then, @sec-result presents the prediction outcome, and @sec-discussion discusses the results and models.

Appendix includes three parts: [Appendix -@sec-appx-pollster] presents methodology overview and evaluation of one certain pollster - Emerson; [Appendix -@sec-appx-survey] provides detailed idealized methodology and survey design for the poll; [Appendix -@sec-model-details] presents additional details during modeling process, including posterior predictive check [Appendix -@sec-ppc], comparison between posterior and prior distributions [Appendix -@sec-pvp], and diagnostics [Appendix -@sec-diagnostics].

@@

## Model Set-up {#sec-model-set}

@@

### Bayesian Model with Spline for Vote Percentage Nationwide

@@

### Bayesian Model with Spline for Vote Percentage and State as Fixed Effect

@@

### Assumptions of the Bayesian Models

@@

## Model Justification {#sec-model-justify}

@@

@@

@@

## Prediction and Spline Fit for Vote Percentage Nationwide

@@

## Prediction and Spline Fit for Vote Percentage by State

@@

## Overview of the Paper {#sec-first-point}

This paper presented a analysis using Bayesian modeling to predict the outcome of the 2024 U.S. Presidential Election. This paper employed the "poll-of-polls" approach to make predictions, which aggregated multiple polls to minimize biases and enhance prediction accuracy.

The voting percentage of Donald Trump is the main response variable, and the number of days since half year before the election "end_date_num" and the regional location "state" are the predictors.

Two distinct Bayesian models were constructed: one that utilized nationwide data, and another that incorporated state-specific effects. The relationship between the response and predictor variables were captured by non-linear natural splines, providing trends of how vote percentage changes as time approaches the election date.

The Bayesian model with natural splines was chosen for analysis because it is appropriate for capturing complex electoral dynamics, allowing for flexibility, evaluating the trend across time, thus providing predictions about the election outcome.

The results investigated by the paper basically met our anticipation. As time approaches the election date, the predicted support for Trump increased stably, increasing from 45% half year ago to nearly 50% recently. Specifically speaking from state perspective, steady increase trend still persists in predictions of supports from people in different states, and several states have high vote percentage lead.

In conclusion, we predict that the voting and support inclination will continue to favor Donald Trump until the day of the election.

## Fluid Nature of Voter Intensions

What we learned from this study is the understanding of how temporal dynamics influence voter preferences in electoral contexts.

Besides providing predictions about the upcoming 2024 election, the upward sloping trend in vote percentage for Donald Trump with respect to days passed since half year ago found by the paper suggests that voters are highly responsive to the political incidence. As mentioned before, frequent presidential debates, promises made by politicians at the last minute, or a sudden outbreak of an event would incite more people to vote for candidate, or change their mind and switch to support the other.

This finding underscores the fluid nature of voter intentions, and the critical role of time in shaping electoral outcomes. For political strategists, the paper emphasizes the importance of imposing more provocative measures as the election date is nearer. For voters, this paper could help them understand the dynamic flow of political elections clearly, and help to maintain distinct, calm overview during election activities.

## State as an Important Factor

Moreover, the second Bayesian model in the paper included state as a fixed effect, providing information on not only the electoral behaviors of the voters from a broad national perspective, but also on the regional variations in support trends.

By considering locational issues, the predictions and analysis about the vote tendency with respect to time could be conducted more completely, and enable people to view the reasons behind the formation of such preferences from the perspective of regional differences. For instance, the model identifies states like South Dakota, where Trump’s support is relatively high, versus states like Massachusetts, which shows a comparatively low, negative coefficient.

The study reveals the disparities in voter support across different locations, reinforcing the idea that a "one-size-fits-all" approach to campaigning is highly inadequate. The finding encourages politicians to encourage politicians to "suit the remedy to the case", and choose different political strategies and means based on the preferences and beliefs of people in various regions respectively, in order to gain the trust and support of voters.

@@

## More to Be Done

Though applying complete statistical modeling and analysis, this study only provided predictions about the election outcome, and trends of vote support to one certain candidate over time. Further researches and investigations are essential to refine the predictive modeling included in this paper.

Firstly, more models could be considered to be used in the predicting procedure. One model type may seem to be too simple and have too many assumptions and limitations, making it difficult to provide an overview of such electoral context. More model validation and sensitivity analyses could be done to improve the modeling process, so that the robustness of the models against different assumptions and priors could be assessed to make the model more fit.

Secondly, data collection methodologies could be improved as well. In this study only one raw dataset was used, arising potential issues in true representativeness of the polling data. Finding and using more data would lead to more accurate and relaible predictions in future electoral analysis.

Last but not least, evaluations on the possible reasons behind the trends of the vote percentage could be investigated. Specifically, since the second Bayesian model included in this paper considered state as a fixed effect, further researches could analysis why such regional variations exist. Is this because people in each region have their own unique sets of values and beliefs, or is it because people in a certain region have experienced the same events? Future studies could explore the integration of additional predictors, and consider various factors such as socio-economic reasons, historical issues, and ethic identity.

@@

## Posterior Predictive Check {#sec-ppc}

During the modeling process, posterior predictive checks (PPC) is used.

PPC is the comparison between what the fitted model predicts and the actual observed data, which validates whether the fitted model is compatible with the observed data. The aim is to detect if the model is inadequate to describe the data [@ppcheck].

@fig-ppcheck shows the posterior predictive checks for both Bayesian models.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheck
#| layout-ncol: 2
#| fig-cap: "Posterior Predictive Check of Two Bayesian Models"
#| fig-subcap: ["Posterior Predictive Check of Bayesian Model 1 (National)", "Posterior Predictive Check of Bayesian Model 2 (by State)"]

pp_check(spline_model_national)

In @fig-ppcheck-1, the black line basically follows the trend of the light blue lines, indicating the observed data consistently falling within the range of simulated data. This suggests that the model provides a good fit, effectively predicting the actual trend of the data.

Similarly, @fig-ppcheck-2 shows the posterior predictive checks for the second model, which predicts Trump's percentage vote over time by state. Although the black line has minor deviation away from the light blue lines, and a part of the black line falls outside the range of simulated lines, the majority part of the black line matches with the light blue lines.

## Posterior vs. Prior {#sec-pvp}

Comparing the posterior with the prior is also necessary for model validation. It examines how the model fits and is affected by the data.

Prior distribution represents the beliefs about the parameter values before observing any data, which is constructed by our initial assumptions. Posterior distribution reflects updated beliefs about the parameters after observing the data.

In the plot, the posterior distribution is demonstrated on the left side, and the prior distribution is on the right side. Each colored dot on the graph represents a parameter, and the horizontal line indicates the uncertainty.

By comparing the posterior with the prior, whether data has a significant impact on the parameter estimates and whether the model is well-fit could be validated.

@fig-model1-posteriorvsprior shows the posterior vs. prior plot for the Bayesian model in the national perspective. We can see that in the posterior distribution, parameters have a smaller spread compared to the prior, indicating less uncertainty. There are also some differences between the prior and posterior distributions, suggesting that the model is well-fit and captures certain patterns in the data.

```{r}
#| echo: false
#| eval: true
#| label: fig-model1-posteriorvsprior
#| fig-cap: "Posterior vs Prior Plot for Model 1 (National)"
#| warning: false

posterior_vs_prior(spline_model_national) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics {#sec-diagnostics}

### R-hat Plots

$\hat{R}$ is a diagnostic measure that compares the variance within each chain to the variance between chains.

R-hat plot shows the values of the $\hat{R}$ statistic on the x-axis, which is a measure of convergence for Markov Chain Monte Carlo (MCMC) simulations in Bayesian analysis, and shows different parameters or chains on the y-axis.

The plot categorizes the parameters based on their $\hat{R}$ values. It is a crucial diagnostic tool in assessing the validity of the model. If most parameter have values $\hat{R} \leq 1.05$, the parameters have good convergence. This indicates that these parameters have likely reached their stationary distribution, and the chains have converged to similar distribution, suggesting that the sampling process has mixed well.

@fig-rhat-1 and @fig-rhat-2 presents R-hat plots for the two Bayesian models constructed in the paper respectively. Both plots show that all parameters in two models have values $\hat{R} < 1.05$ and around 1, indicating good convergence and a well-fitted model.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-rhat
#| layout-ncol: 2
#| fig-cap: "R-hat Plots of Two Bayesian Models"
#| fig-subcap: ["R-hat Plot of Bayesian Model 1 (National)", "R-hat Plot of Bayesian Model 2 (by State)"]

plot(spline_model_national, "rhat")

plot(spline_model_state, "rhat")
```

### Trace Plots

Trace plot shows samples from all the chains. Each chain is represented by a different color or line style.

If all chains appear to mix well and oscillate around the same mean, and overlap significantly, it suggests that the parameter estimates are stable.

@fig-model1-trace presents a trace plot for the Bayesian model which is in the national perspective. We can see that there is no clear separation between the chains, suggesting that there is lack of convergence issue and well-fit model. However, relatively large fluctuations in the plot indicate some uncertainty, and less effective and precise outcome. This could be probably fixed by adding more data into the modeling procedure.

```{r}
#| echo: false
#| eval: true
#| label: fig-model1-trace
#| fig-cap: "Trace Plot for Model 1 (National)"
#| warning: false

plot(spline_model_national, "trace")
```
